---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## 1. Global Settings and libraries
```{r global_options}
library(knitr)
opts_chunk$set(fig.path='figure/')
library(dplyr)
library(ggplot2)
```

## 2. Loading and preprocessing the data
1. Unzip the data file activity.zip and read the data into a data frame named activity
```{r unzip}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

## 3. What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day ignoring the missing values
```{r meantotal}
numSteps <- activity[complete.cases(activity),] %>%
        group_by(date) %>%
        summarize_each(funs(sum)) %>% 
        select(date, steps)
```

2. Make a histogram of the total number of steps taken each day
```{r tothist}
hist(numSteps$steps, 
     breaks = floor(nrow(numSteps)/2), 
     main = "Histogram of Total numbers of Steps taken each day", 
     xlab = "Steps per Day", 
     col = "blue")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r meanmed}
meanSteps <- round(mean(numSteps$steps))
medianSteps <- median(numSteps$steps)

cat("Mean Steps per day   : ",meanSteps,"\nMedian Steps per day : ",medianSteps)
```



## 4. What is the average daily activity pattern?
```{r avgint}
intPatternMean <- activity[complete.cases(activity),] %>% 
        group_by(interval) %>% 
        select(interval, steps) %>% 
        summarize_each(funs(mean))

with(intPatternMean, plot(steps ~ interval, 
                          type = "l", 
                          ylab = "Average number of steps taken",
                          xlab = "5 minute interval",
                          main = "Daily Activity Pattern",
                          col  = "blue")
     )
```



## 5. Imputing missing values

1. Calculate and report the total number of missing values in the dataset
```{r miss1}
missValCount <- dim(activity[!complete.cases(activity),])[1]
cat("Total number of missing values in the dataset is : ",missValCount)
```

2. Create a new dataset from the original dataset but with the missing values for steps filled in with the mean of the 5-minute interval across all days rounded off
```{r newds}
## Duplicate the original data frame
activity1 <- activity

## Fill NA values in activity1 with median of 5 minute intervals
for (i in 1:dim(intPatternMean)[1]) {
        s1 <- c(round(intPatternMean[i, "steps"]))
        int1 <- c(intPatternMean[i, "interval"])
        activity1[activity1$interval == int1 
                  & is.na(activity1$steps),"steps"] <- s1
}
```

3. Make a histogram of the total number of steps taken each day
```{r meantotal1}
numSteps1 <- activity1 %>%
        group_by(date) %>%
        summarize_each(funs(sum)) %>% 
        select(date, steps)

hist(numSteps1$steps, 
     breaks = floor(nrow(numSteps)/2), 
     main = "Revised Total numbers of Steps taken each day", 
     xlab = "Steps per Day", 
     col = "blue")
```

4. Calculate and report the mean and median total number of steps taken per day
```{r meanmed1}
meanSteps1 <- round(mean(numSteps1$steps))
medianSteps1 <- median(numSteps1$steps)
cat("Revised Mean and Median Values:","\nMean Steps per day   : ",meanSteps1,
    "\nMedian Steps per day : ",medianSteps1)
```

5. After supplying missing values, the mean of the steps remain unchanged while the median changed from `r format(medianSteps, scientific=F)` to `r format(medianSteps1, scientific=F)`. Observed impact of the imputing missing values as per the above methodlogy only resulted in this very minor change in the median. 



## 6. Are there differences in activity patterns between weekdays and weekends?
```{r weekend}
## Create a day type factor variable from the weekday
daytype <- weekdays(as.Date(activity1$date))
daytype <- gsub("Sunday|Saturday", "Weekend", daytype)
daytype <- gsub("Monday|Tuesday|Wednesday|Thursday|Friday", "Weekday", daytype)
daytype <- as.factor(daytype)

## Bind the new variable to the df - create a new df by summarizing means steps
## over interval and daytype
activity2 <- cbind(activity1,daytype)  %>% 
        group_by(interval,daytype) %>% 
        summarize_each(funs(mean), steps)

## Plot using ggplot system
g1 <- ggplot(activity2, aes(interval, steps))

g1 + geom_line(col="blue") + 
        facet_grid(daytype ~ .) + 
        theme_linedraw(base_size = 12) + 
        labs(list(title = "Activity Patterns - Weekday vs Weekend", 
                  x = "Time Interval", 
                  y = "# of Steps")
             )
```
